Making Sequential Programs Deterministic

### 2019-03-28 design.tex
Edits in section 2.0
Edits in section 2.1

\section{Design}\label{design}

%% This section is too basic, and therefore is boring because most readers
%% will already know it.  It also delays the interesting parts of the paper
%% for too long.  We can introduce concepts as they are needed in the rest
%% of the paper.
% \subsection{Type checking background}\label{type-checking}
% The major components of a type system include: 1) \textit{types}, 2) \textit{subtyping rules}, and 3) \textit{dataflow analysis}.
% A \textit{type} serves as an abstraction for the set of acceptable values for any expression. The types in a type system form a lattice of finite height. 
% The hierarchy of types in this lattice defines subtyping relationships among them.
% In our framework, we require every type hierarchy to define a unique\<@Top> and a \<@Bottom> element. This ensures that
% any given pair of types has a \textit{least upper bound} and a \textit{greatest lower bound}.
% Throughout this paper, we use the notation \<@A> <: \<@B> to denote that type \<@A> is a subtype of type \<@B>.
% As an illustration of the type hierarchy, consider the lattice of types shown in Figure~\ref{fig-example-lattice}.
% \begin{figure}
%     \begin{center}
%         \includegraphics[scale=0.15]{lattice}
%     \end{center}
%     \caption{Example type hierarchy}
%     \label{fig-example-lattice}
% \end{figure}
% In this type system, \<@C> is a subtype of \<@A> as well as \<@B> and types \<@A> and \<@B> are incomparable.
% The type checker now performs additional type checking (similar to javac's type checking) with respect to this type hierarchy
% and reports any violations. The snippet in Figure~\ref{code-invalid1} would result in an error at the assignment statements:
% \begin{figure}
%     \begin{verbatim}
%     @A int x; @B int y; @C int z;
%     x = y;    // Because types @A and @B are incomparable.
%     z = x;    // Because @C is a subtype of @A.
%     \end{verbatim}
% \caption{Example: Invalid assingment}
% \label{code-invalid1}
% \end{figure}
% 
% %Method overriding and example
% For method overriding, the usual rules apply for parameters and arguments i.e covariant subtyping for parameters
% and contravariant subtyping for return types. Consider the example in Figure~\ref{code-invalid2} where method \<foo> of \<class X> is overridden
% in \<class Y>.
% \begin{figure}
%     \begin{verbatim}
%     class X {
%         @B int foo(@C int param) {}
%     }
%     class Y extends X {
%         @C int foo(@B int param) {}
%     }
%     \end{verbatim}
%     \caption{Example: Invalid Method override}
%     \label{code-invalid2}
% \end{figure}
% 
% The overriding method is invalid for two reasons: type of the parameter (\<@B>) is not a subtype of the parameter type in the overridden method (\<@C>) and the return type (\<@C>) is not a supertype of the return type of the overridden method (\<@B>).
% 
% %collections type parameters - invariant
% Collection types are invariant. So, if a \<List> x is declared as \codeid{@A List<@A String> x;} and another \<List y> is declared
% as  \codeid{@A List<@B String> y;}, writing \<x = y> would result in an error.
% 
% %covariant array types
% Arrays follow covariant subtyping rules in java. Therefore, the assignment statement \<@A int @A[] x = y;> where \<y> was 
% declared as \<@B int @A[] y> would type check without any warning.

@@

In a type system,
a \textit{type} represents a set of values.  It abstracts or restricts the
set of possible run-time values that an expression may evaluate to.
A programming language provides \emph{basetypes}, such as \<int>.
A \textit{type qualifier} on a \textit{type} adds additional constraints;
that is, it reduces the size of the set of values.
In Java, a type qualifier is written as a type annotation (which has a leading \<@>), as in
\<@Positive int>.
A type checker can verify the correctness of a \textit{type} via subtyping
rules and dataflow analysis.

@@

    \caption{Type checking rules.  $\sqsubseteq$ is overloaded for classes,
    basetypes, type qualifiers, and types.}
    \label{typecheck-rules}
\end{figure}

@@

\begin{wrapfigure}{R}{0.3\textwidth}
    \begin{center}
        \includegraphics[scale=0.5]{determinism}
    \end{center}
    \caption{Determinism type qualifier hierarchy}
    \label{fig-determinism-hierarchy}
\todo{Drop the at-signs (@).  The text is weirdly small for the box.
  Probably just redraw this.}
\end{wrapfigure}

The core of the determinism type system is the following type qualifiers (see Figure~\ref{fig-determinism-hierarchy}):
%\todo{introduce just @Det and @NonDet first, and then @OrderNonDet after polymorphism.}
\begin{itemize}
    \item \<NonDet> indicates
    that the expression might have different values in two different executions.
    \item \<OrderNonDet> indicates that the expression is a collection or
      map that contains the same elements in every execution, but possibly
      in a different order.
    \item \<Det> indicates that
    the expression evaluates to the same value (with respect to \<.equals()>) in all
    executions; for a collection, iteration also yields the values in the same
    order.
\end{itemize}

@@

In Figure~\ref{code-determinism}, we present some of the JDK methods that we have annotated with our determinism types and give examples of
client code that would produce errors at compile time.
\begin{figure}
    \begin{verbatim}
    // Annotated JDK methods
    public class Random implements java.io.Serializable {
        public @NonDet Random() {}
    }
    public class PrintStream extends FilterOutputStream 
        implements Appendable, Closeable {
        public void println(@Det Object x) {}
    }

@@

Figure~\ref{type-validity} shows the type validity\todo{The caption says
  ``type well formedness''.  Choose one term and use it consistently
  throughout.}
rules of our type system. A type $\tau$ is represented as $\kappa \ \beta$
where $\kappa$ is a type qualifier (that is, \<NonDet>, \<OrderNonDet>, or
\<Det>) and $\beta$ is a Java basetype.
As stated in the rule $\textsc{any type}$, the types \<Det> and \<NonDet>
can be written on any Java basetype.\todo{As noted in figure caption, that's
  not true, in some contexts such as a type argument.}

The determinism type qualifier on a collection must be a supertype or equal to
its element type qualifier (Rule $\textsc{ordernondet collection}$ in Figure~\ref{fig-determinism-collections}).
If the Collection is \<NonDet>, then the type argument may not be
\<Det> or \<OrderNonDet> (Rule $\textsc{nondet collection}$ in Figure~\ref{type-validity}). Although such types could exist, they are
disallowed to prevent the following bug:

@@

Figure~\ref{fig-poly-resolution} presents the rule for type checking of methods annotated with \<PolyDet> or its variants.
A polymorphic method type checks if every instantiation of \<PolyDet> type checks. If this method has \<PolyDet>$\uparrow$ or \<PolyDet>$\downarrow$
in its signature then the instantiation of \<PolyDet> at that location gets replaced by the appropriate type as shown in Table~\ref{tab-poly-resolutions}.

@@

\TheDeterminismChecker has additional type refinement rules for the sorting
methods such as \<List.sort>, \<Collections.sort>, and \<Arrays.sort>:
 as shown in Figure~\ref{fig-sorting}. 
\begin{figure}
%     $\infer[\textsc{list sort}]{\Gamma \vdash x : \|Det \ List|\angles{\|Det| \ \beta}}{\Gamma \vdash x : \|OrderNonDet \ List|\angles{\|Det| \ \beta} & \|x.sort()|}$
%     
%     \bigskip

@@

    \caption{Sorting type refinement rules}
    \label{fig-sorting}
\end{figure}
\TheDeterminismChecker type refines a receiver of type \codeid{OrderNonDet List} to a \codeid{Det List} if all the
type argument of this \codeid{List} have the type qualifier \codeid{Det} (Rule LIST SORT in Figure~\ref{fig-sorting}).
As per these rules, if \codeid{sort()} is called on a receiver of type \codeid{OrderNonDet List<Det Set<Det String>>}, it will
be type refined to \codeid{Det List<Det Set<Det String>>}. However a receiver of type \codeid{OrderNonDet List<OrderNonDet Set<Det String>>} will not be type refined.

@@

We prohibit the creation of a \codeid{Det HashMap} or an \codeid{OrderNonDet TreeMap} in our determinism checker.
Figure~\ref{fig-creation-rules} formalizes these rules. Similar constraints apply to \<HashSet>s and \<TreeSet>s which are
also expressed by the last two rules of Figure~\ref{fig-creation-rules}.

@@

\subsection{Determinism checker specific rules for precision}\label{precision}
We annotated the \codeid{equals()} method of the \<Set> class in the JDK as
\begin{verbatim}
PolyDet$\uparrow$ boolean equals(PolyDet Set<E> this, PolyDet Object o)
\end{verbatim}
While this is sound, it is imprecise in the case of \codeid{OrderNonDet Sets}. We added the following rule shown in Figure~\ref{fig-precision-rules} to our checker to improve precision.
\begin{figure}
    \todo{add line break}
    $\infer[\textsc{set precision}]{\Gamma \vdash \|z : Det|}{\Gamma \vdash \|x| : \|OrderNonDet| \ \|Set|\angles{ \kappa_{x} \ \beta_{x}}, \|y| : \|OrderNonDet| \ \|Set|\angles{ \kappa_{y} \ \beta_{y}} & \|z = x.equals(y)| & \angles{ \kappa_{x} \ \beta_{x}} \nsqsubseteq \angles{ \|OrderNonDet \ List| } & \angles{ \kappa_{y} \ \beta_{y}} \nsqsubseteq \angles{ \|OrderNonDet \ List| } }$

@@

Similarly, the \codeid{get()} method in the \<Map> class is annotated with polymorphic qualifiers and the rule $\sc{\|MAP \ PRECISION|}$ in Figure~\ref{fig-precision-rules} improves precision for \<OrderNonDet> receivers.

>>>>

\section{A type system for determinism}\label{design}

This section presents the key aspects of the design of our type system,
called \ourTypeSystem, in the context of a core calculus for an
object-oriented language.

\subsection{Preliminaries}

%% This section is too basic, and therefore is boring because most readers
%% will already know it.  It also delays the interesting parts of the paper
%% for too long.  We can introduce concepts as they are needed in the rest
%% of the paper.
% \subsection{Type checking background}\label{type-checking}
% The major components of a type system include: 1) \textit{types}, 2) \textit{subtyping rules}, and 3) \textit{dataflow analysis}.
% A \textit{type} serves as an abstraction for the set of acceptable values for any expression. The types in a type system form a lattice of finite height. 
% The hierarchy of types in this lattice defines subtyping relationships among them.
% In our framework, we require every type hierarchy to define a unique\<@Top> and a \<@Bottom> element. This ensures that
% any given pair of types has a \textit{least upper bound} and a \textit{greatest lower bound}.
% Throughout this paper, we use the notation \<@A> <: \<@B> to denote that type \<@A> is a subtype of type \<@B>.
% As an illustration of the type hierarchy, consider the lattice of types shown in \cref{fig-example-lattice}.
% \begin{figure}
%     \begin{center}
%         \includegraphics[scale=0.15]{lattice}
%     \end{center}
%     \caption{Example type hierarchy}
%     \label{fig-example-lattice}
% \end{figure}
% In this type system, \<@C> is a subtype of \<@A> as well as \<@B> and types \<@A> and \<@B> are incomparable.
% The type checker now performs additional type checking (similar to javac's type checking) with respect to this type hierarchy
% and reports any violations. The snippet in \cref{code-invalid1} would result in an error at the assignment statements:
% \begin{figure}
%     \begin{verbatim}
%     @A int x; @B int y; @C int z;
%     x = y;    // Because types @A and @B are incomparable.
%     z = x;    // Because @C is a subtype of @A.
%     \end{verbatim}
% \caption{Example: Invalid assingment}
% \label{code-invalid1}
% \end{figure}
% 
% %Method overriding and example
% For method overriding, the usual rules apply for parameters and arguments i.e covariant subtyping for parameters
% and contravariant subtyping for return types. Consider the example in \cref{code-invalid2} where method \<foo> of \<class X> is overridden
% in \<class Y>.
% \begin{figure}
%     \begin{verbatim}
%     class X {
%         @B int foo(@C int param) {}
%     }
%     class Y extends X {
%         @C int foo(@B int param) {}
%     }
%     \end{verbatim}
%     \caption{Example: Invalid Method override}
%     \label{code-invalid2}
% \end{figure}
% 
% The overriding method is invalid for two reasons: type of the parameter (\<@B>) is not a subtype of the parameter type in the overridden method (\<@C>) and the return type (\<@C>) is not a supertype of the return type of the overridden method (\<@B>).
% 
% %collections type parameters - invariant
% Collection types are invariant. So, if a \<List> x is declared as \codeid{@A List<@A String> x;} and another \<List y> is declared
% as  \codeid{@A List<@B String> y;}, writing \<x = y> would result in an error.
% 
% %covariant array types
% Arrays follow covariant subtyping rules in java. Therefore, the assignment statement \<@A int @A[] x = y;> where \<y> was 
% declared as \<@B int @A[] y> would type check without any warning.

@@

\todo{Greek letters go after this paragraph.}

In a type system,
a \textit{type} represents a set of values.  It abstracts or restricts the
set of possible run-time values that an expression may evaluate to.
A programming language provides \emph{basetypes}, such as \<int>.
A \textit{type qualifier} on a basetype adds additional constraints;
that is, it reduces the size of the set of values.
An example type qualifier is \<Positive>,\todo{should all type qualifiers
  be lowercase in this section??} and a type (which combines a qualifier
anda basetype) is \<Positive int>.
A polymorphic type such as \<List> can be instantiated by a type argument,
as in \ptype{List}{Positive int}.

A type checker verifies the types written in a program.

@@

    \caption{Type checking rules.  $\sqsubseteq$ is overloaded for classes,
    basetypes, type qualifiers, and types.}
    \label{fig:typecheck-rules}
\end{figure}

@@

\begin{wrapfigure}{R}{0.3\textwidth}
    \begin{center}
        \includegraphics[scale=0.5]{determinism}
    \end{center}
    \caption{Determinism type qualifier hierarchy}
    \label{fig:determinism-hierarchy}
\todo{Drop the at-signs (@).  The text is weirdly small for the box.
  Probably just redraw this.}
\end{wrapfigure}

The core of the determinism type system is the following type qualifiers:
%\todo{introduce just @Det and @NonDet first, and then @OrderNonDet after polymorphism.}
\begin{itemize}
    \item \<NonDet> indicates
    that the expression might have different values in two different executions.
    \item \<OrderNonDet> indicates that the expression is a collection or
      map that contains the same elements in every execution, but possibly
      in a different order.
    \item \<Det> indicates that
    the expression evaluates to the same value (with respect to \<.equals()>) in all
    executions; for a collection, iteration also yields the values in the same
    order.
\end{itemize}
\Cref{fig:typecheck-rules,fig:determinism-hierarchy} show their subtyping relationship.

@@

In \cref{code-determinism}, we present some of the JDK methods that we have annotated with our determinism types and give examples of
client code that would produce errors at compile time.
\begin{figure}
    \begin{verbatim}
    // Annotated JDK methods
    public class Random implements java.io.Serializable {
        public @NonDet Random() {}
    }
    public class PrintStream extends FilterOutputStream 
        implements Appendable, Closeable {
        public void println(@Det Object x) {}
    }

@@

\Cref{type-validity} shows the type validity\todo{The caption says
  ``type well formedness''.  Choose one term and use it consistently
  throughout.}
rules of our type system. A type $\tau$ is represented as $\kappa \ \beta$
where $\kappa$ is a type qualifier (that is, \<NonDet>, \<OrderNonDet>, or
\<Det>) and $\beta$ is a Java basetype.
As stated in the rule $\textsc{any type}$, the types \<Det> and \<NonDet>
can be written on any Java basetype.\todo{As noted in figure caption, that's
  not true, in some contexts such as a type argument.}

The determinism type qualifier on a collection must be a supertype or equal to
its element type qualifier (Rule $\textsc{ordernondet collection}$ in \cref{fig-determinism-collections}).
If the Collection is \<NonDet>, then the type argument may not be
\<Det> or \<OrderNonDet> (Rule $\textsc{nondet collection}$ in \cref{type-validity}). Although such types could exist, they are
disallowed to prevent the following bug:

@@

\Cref{fig-poly-resolution} presents the rule for type checking of methods annotated with \<PolyDet> or its variants.
A polymorphic method type checks if every instantiation of \<PolyDet> type checks. If this method has \<PolyDet>$\uparrow$ or \<PolyDet>$\downarrow$
in its signature then the instantiation of \<PolyDet> at that location gets replaced by the appropriate type as shown in Table~\ref{tab-poly-resolutions}.

@@

\TheDeterminismChecker has additional type refinement rules for the sorting
methods such as \<List.sort>, \<Collections.sort>, and \<Arrays.sort>:
 as shown in \cref{fig-sorting}.
\begin{figure}
%     $\infer[\textsc{list sort}]{\Gamma \vdash x : \|Det \ List|\angles{\|Det| \ \beta}}{\Gamma \vdash x : \|OrderNonDet \ List|\angles{\|Det| \ \beta} & \|x.sort()|}$
%     
%     \bigskip

@@

    \caption{Sorting type refinement rules}
    \label{fig-sorting}
\end{figure}
\TheDeterminismChecker type refines a receiver of type \codeid{OrderNonDet List} to a \codeid{Det List} if all the
type argument of this \codeid{List} have the type qualifier \codeid{Det} (Rule LIST SORT in \cref{fig-sorting}).
As per these rules, if \codeid{sort()} is called on a receiver of type \codeid{OrderNonDet List<Det Set<Det String>>}, it will
be type refined to \codeid{Det List<Det Set<Det String>>}. However a receiver of type \codeid{OrderNonDet List<OrderNonDet Set<Det String>>} will not be type refined.

@@

We prohibit the creation of a \codeid{Det HashMap} or an \codeid{OrderNonDet TreeMap} in our determinism checker.
\Cref{fig-creation-rules} formalizes these rules. Similar constraints apply to \<HashSet>s and \<TreeSet>s which are
also expressed by the last two rules of \cref{fig-creation-rules}.

@@

\subsection{Determinism checker specific rules for precision}\label{precision}
We annotated the \codeid{equals()} method of the \<Set> class in the JDK as
\begin{verbatim}
PolyDet$\uparrow$ boolean equals(PolyDet Set<E> this, PolyDet Object o)
\end{verbatim}
While this is sound, it is imprecise in the case of \codeid{OrderNonDet Sets}. We added the following rule shown in \cref{fig-precision-rules} to our checker to improve precision.
\begin{figure}
    \todo{add line break}
    $\infer[\textsc{set precision}]{\Gamma \vdash \|z : Det|}{\Gamma \vdash \|x| : \|OrderNonDet| \ \|Set|\angles{ \kappa_{x} \ \beta_{x}}, \|y| : \|OrderNonDet| \ \|Set|\angles{ \kappa_{y} \ \beta_{y}} & \|z = x.equals(y)| & \angles{ \kappa_{x} \ \beta_{x}} \nsqsubseteq \angles{ \|OrderNonDet \ List| } & \angles{ \kappa_{y} \ \beta_{y}} \nsqsubseteq \angles{ \|OrderNonDet \ List| } }$

@@

Similarly, the \codeid{get()} method in the \<Map> class is annotated with polymorphic qualifiers and the rule $\sc{\|MAP \ PRECISION|}$ in \cref{fig-precision-rules} improves precision for \<OrderNonDet> receivers.
